Leadership across cultures and contexts: an exploratory analysis

### 2020-04-03 Response to reviewers.Rmd

**Finally, the reviewers made a number of valuable suggestions, not enumerated above,  that should help you sharpen you presentation and discussion. Please attend to those.**

## Reviewer 1 comments

**This manuscript presents a descriptive analysis of the HRAF to provide information about leadership in traditional societies. Overall, I enjoyed reading the paper and thought it was a useful resource. It provides a comprehensive analysis of leadership across a wide swath of traditional cultures. It reports on the types of functions leaders serve, the qualities possessed by leaders, and the costs and benefits of leadership experienced by both leaders and followers. The paper will likely be of interest to anyone working in the area of leadership, social hierarchy, or group behavior. I provide some comments below aimed at helping the authors increase the potential accessibility and impact of the work.**

We greatly appreciate the detailed comments and very useful suggestions.

@@

We added a few additional details in the methods section on the eHRAF and ethngraphic sample and now more clearly direct readers to the original pubication from which the database was built, which includes additional details.

We also include additional details in the results and supplimentary information on the corpus of ethnographic documents which provide the sample of ethnographic paragraphs. NEED TO DO.

**I don't work with ethnographic data, so perhaps this question is naïve, but: What conclusions can we draw about the actual frequency of leadership functions and traits from the frequency with which they are mentioned in the ethnographic record? There is an implicit assumption that the two are closely aligned. Has such an alignment been confirmed? Is that even testable?**

The reviewer brings up an important point. For one, this is partly our motivation for representing the percentage of cultures which provide at least one text record supporting each of the operationalized variables. Despite this low threshold for accepting supporting evidence, the ethnogrtaphic record (sampled here) does not provide support for every variable from every culture. At this culture-level we feel more confident our data provide a estiamte of cross-cultural frequency, albeit an inherently noisy, "course-grained" estimation. Also, the entire field of ethnology (the branch of anthropology focused on comparing charachteristics of populations across the globe) is based on developing comparable and often quantitaive measures from primary ethnogrpahic information. For example, the Ethnographic Atlas, which provides hundreds of coded quantitaive and qualitive variables on thousands of cultures and the Standard Cross-Cultural Sample, which provides thousands of coded quantitative and qualitative variables on 186 cultures, were developed in essentially the same way we have developed our operatiopnalized variables here. Developing and contrasting coded measures from primary ethnographic documents is not without limitations, but does represent (perhaps) the only method to systematically and empirically compare measures across a large sample of cultures/populations. At the text-record level drawing strong conclusions about the actual frequency of traits is slightly more problematic. However, fundamentally these results provide a systematic view of the ethnographic record on these measures. From our results field researchers and theoreticians can draw inference from a more comprehensive view of human cultural diversity rather than haphazardly self-selecting a few ethngoraphic cases studies.

@@

**The Wofford study (lines 214-233) and GLOBE study reminds me of the vast array of descriptive models in leadership/managerial studies- transactional vs. transformational leadership and their sub-components being one of the most popular. So what should a reader make of this literature, with respect to evolution of leadership in small-scale societies? Do bureaucracies, formal legal systems, formal education-based skills, and large inequalities in remuneration limit universality of descriptive models from leadership/managerial studies? I'd be more explicit on this in the introduction. For the discussion section: how does your own descriptive modeling of leadership in traditional societies agree or not with Wofford or other models from large-scale societies?**

**There is little mention of gender in the Qualities of leadership section, even though this might be first attribute an alien might recognize as covarying with community-level leadership in human societies. Even in the more egalitarian societies, men often wield more overt political influence during camp or band-wide political discussions. Without getting into arguments about patriarchy and its origins, it might at least be useful to cite your own work on Chabu and work on the Tsimane (von Rueden et al. 2018) that finds similar traits predict leadership for men and women, implying its not gender per se that gives men a leadership advantage in politics.**

We now discuss sex-differences in community and political leadership, and refernce in greater deatils (albeit breif) the universal male bias in community leadership in the introuction and cite these works.

### 2020-04-03 leadership_across_cultures_contexts.Rmd
Adding text to main MS.
Added stan_glmer models.

Across culturally and economically diverse populations there are some broad commonalities in the qualities of community leaders. When strong community leaders emerge, they are most typically high status males who embody the ideal qualities most respected by the group and necessary for successful leadership in the domains in which they have influence [@henrich_evolution_2001; @van_vugt_evolutionary_2006; @collier_politics_1981; @low_men_1992; @roscoe_social_2019]. Elders, respected individuals, and highly intelligent or knowledgeable individuals tend to have greater influence over community decisions across cultures, including among the most egalitarian hunter-gatherers, stratified chiefdoms, and contemporary state-level organizations [@silverman_how_1978; @van_vugt_cognitive_2007; @tooby_cognitive_2006; @garfield_review_2019; @bass_handbook_1990]. Drawing on cross-national data through the GLOBE project from over 15,000 middle managers across 60 countries, @den_hartog_culture_1999 conclude that elements of charismatic and transformational leadership -- leadership styles when leaders express an inclusive vision of future progress, lead through inspiration, and develop initiatives in collaboration -- are universally endorsed across nations and are qualities perceived to be associated with effective leadership. In particular, inducing motivation, having foresight, being encouraging and communicative, trustworthiness, and building self-confidence among followers are suggested to be universal attributes of outstanding leadership in managerial contexts.

@@

We estimate the cross-cultural frequency in the functions and qualities of leaders and the costs and benefits of leadership for both leaders and followers using an ethnographic database previously built and published (doi.org/10.5281/zenodo.2541999), from which @garfield_evolutionary_2019 "tested" three theoretical models against the ethnographic record. The electronic Human Relations Area Files (eHRAF) is a digital collection of ethnographies, subject-coded at teh paragraph level following a the Outline of Cultural Materials (OCM) coding scheme. The eHRAF is a subset of the larger HRAF and is continually growing in scope.

@@

The operationalized qualities and functions (See Table \@ref(tab:qualfuncvars)) were developed from the ethnographic text records in the database. First ZG read a subset of text records and developed initial variable lists and operationalizations of the qualities and functions in those text records. Then, KS coded all 1212 text records (in a random order) for the operationalized qualities, functions, cost, and benefits, while also creating new variables and operationalizations for additional qualities and functions encountered. After coding half of the random sample KS and ZG revised the operational definitions to produce the final list of variables in Table \@ref(tab:qualfuncvars). ZG then followed the same process of coding the entire sample of 1212 text records for the `r length(c(quality_vars, function_vars, leader_cost_vars, leader_benefit_vars, follower_benefit_vars, follower_cost_vars))` operationalized variables, as well as a measure of group structure coded as: *residential subgroup*, which includes informal groups of co-residents, age-based groups, performance groups; *kin group*, which includes lineages, phraties, and other kin-based organizations; *economic group*, which includes subsistence groups, market groups, and other groups with primarily economic goals; *military group*, which includes all groups related to inter-group conflcit; *religious group*, which includes groups collected for spiritual or supernatural purpose; *political group (community)*, which includes political groups at the level fo the community, i.e., the political leader can potentially directly interact with most community members such as most village headmen, or *political group (supracommunity)*, which includes political groups that encompass multiple residential communities, such as complex chiefdoms, regional political leaders, and kings or state-level leaders. ZG and KS then discussed every instance of disagreement between their respective coding to produce a consensus matrix, which is used in the analyses here and added to the published data repository (a data frame named *leader_text2*).

@@

We explored whether document features, such as date of publication and gender of authors, influenced reporting of leader functions and qualities. There were `r tab_author_gender['male']` male authors, `r tab_author_gender['female']` females authors, and `r tab_author_gender['unkown']` authors whose gender we could not determine; some of these authors were co-authors. We then tested if text records with a female author or co-author were more likely to discuss female leaders. Of the `r sum(tab_coauthor['FALSE',])` text records with no female co-author, `r tab_coauthor['FALSE','yes']` (`r 100*signif(tab_coauthor['FALSE','yes']/sum(tab_coauthor['FALSE',]),2)`%) reported female leadership. Of the `r sum(tab_coauthor['TRUE',])` text records with female co-author, on the other hand, `r tab_coauthor['TRUE','yes']` (`r 100*signif(tab_coauthor['TRUE','yes']/sum(tab_coauthor['TRUE',]),2)`%) reported female leadership, suggesting that female ethnographers were more likely to report female leaders. However, after accounting for the non-independence of text records (i.e., multiple records from the same document) using a logistic mixed effects regression model, we found that although there was a positive effect of co-author gender on reporting female leadership (OR = `r signif(mm_coauthorOR, 2)`), it was not statistically significant (p = `r signif(mm_coauthorP, 2)`).

>>>>

There are some broad commonalities in the qualities of community leaders across culturally and economically diverse populations. When strong community leaders emerge, they are most typically high status males who embody the ideal qualities most respected by the group and necessary for successful leadership in the domains in which they have influence [@henrich_evolution_2001; @van_vugt_evolutionary_2006; @collier_politics_1981; @low_men_1992; @roscoe_social_2019]. Substantial evidence suggests there is a near universal male-bias in community, political, or group-level leadership across human societies [@low_sex_1992; @whyte_status_1978; @sanday_female_1981], as well as across diverse non-human social species [@smith_obstacles_2018; @tiger_female_1978]. @low_sex_1992 found in a systematic cross-cultural sample that in only seven of 93 societies political leaders of both sexes were present and in only two societies were female political leaders identified as equally powerful as their male counterparts. Even among these rare cases, powerful female political leaders are still infrequent within the group. Women were more likely to hold influential leadership positions within kin groups [@low_sex_1992]. More recently, empirical studies have suggested sex-differences in political influence and leadership across human societies including among more egalitarian cultural contexts are more likely driven by ecological and social structural factors, rather than sex (or gender identity) per se, including social capital and access to same-sex allies [@yanca_female_2004] and/or sex differences in the sexual division of labor [@von_rueden_sex_2018]. Ethnographic fieldwork comparing leaders and non-leaders in two distinct small-scale societies has revealed sex is not a strong predictor of community influence or elected leadership positions, after accounting for a broad range of demographic, phenotypic, cognitive, social, and other traits derived from the evolutionary theoretical literature [e.g., @garfield_investigating_2019; @von_rueden_sex_2018].

Across cultural contexts elders, respected individuals, and highly intelligent or knowledgeable individuals tend to have greater influence over community decisions across cultures, including among the most egalitarian hunter-gatherers, stratified chiefdoms, and contemporary state-level organizations [@silverman_how_1978; @van_vugt_cognitive_2007; @tooby_cognitive_2006; @garfield_review_2019; @bass_handbook_1990]. Drawing on cross-national data through the GLOBE project from over 15,000 middle managers across 60 countries, @den_hartog_culture_1999 conclude that elements of charismatic and transformational leadership -- leadership styles when leaders express an inclusive vision of future progress, lead through inspiration, and develop initiatives in collaboration -- are universally endorsed across nations and are qualities perceived to be associated with effective leadership. In particular, inducing motivation, having foresight, being encouraging and communicative, trustworthiness, and building self-confidence among followers are suggested to be universal attributes of outstanding leadership in managerial contexts.

@@

We estimate the cross-cultural frequency in the functions and qualities of leaders and the costs and benefits of leadership for both leaders and followers using an ethnographic database previously built and published (doi.org/10.5281/zenodo.2541999), from which @garfield_evolutionary_2019 "tested" three theoretical models against the ethnographic record. The electronic Human Relations Area Files (eHRAF) is a digital collection of ethnographies, subject-coded at the paragraph level following a the Outline of Cultural Materials (OCM) coding scheme. The eHRAF is a subset of the larger HRAF and is continually growing in scope.

@@

The operationalized qualities and functions (See Table \@ref(tab:qualfuncvars)) were developed from the ethnographic text records in the database. First ZG read a subset of text records and developed initial variable lists and operationalizations of the qualities and functions in those text records. Then, KS coded all 1212 text records (in a random order) for the operationalized qualities, functions, cost, and benefits, while also creating new variables and operationalizations for additional qualities and functions encountered. After coding half of the random sample KS and ZG revised the operational definitions to produce the final list of variables in Table \@ref(tab:qualfuncvars). ZG then followed the same process of coding the entire sample of 1212 text records for the `r length(c(quality_vars, function_vars, leader_cost_vars, leader_benefit_vars, follower_benefit_vars, follower_cost_vars))` operationalized variables, as well as a measure of group structure coded as: *residential subgroup*, which includes informal groups of co-residents, age-based groups, performance groups; *kin group*, which includes lineages, phratries, and other kin-based organizations; *economic group*, which includes subsistence groups, market groups, and other groups with primarily economic goals; *military group*, which includes all groups related to inter-group conflict; *religious group*, which includes groups collected for spiritual or supernatural purpose; *political group (community)*, which includes political groups at the level fo the community, i.e., the political leader can potentially directly interact with most community members such as most village headmen, or *political group (supracommunity)*, which includes political groups that encompass multiple residential communities, such as complex chiefdoms, regional political leaders, and kings or state-level leaders. ZG and KS then discussed every instance of disagreement between their respective coding to produce a consensus matrix, which is used in the analyses here and added to the published data repository (a data frame named *leader_text2*).

@@

We explored whether document features, such as date of publication and gender of authors, influenced reporting of leader functions and qualities. There were `r tab_author_gender['male']` male authors, `r tab_author_gender['female']` females authors, and `r tab_author_gender['unkown']` authors whose gender we could not determine; some of these authors were co-authors. We then tested if text records with a female author or co-author were more likely to discuss female leaders. Of the `r sum(tab_coauthor['FALSE',])` text records with no female co-author, `r tab_coauthor['FALSE','yes']` (`r 100*signif(tab_coauthor['FALSE','yes']/sum(tab_coauthor['FALSE',]),2)`%) reported female leadership. Of the `r sum(tab_coauthor['TRUE',])` text records with a female co-author, on the other hand, `r tab_coauthor['TRUE','yes']` (`r 100*signif(tab_coauthor['TRUE','yes']/sum(tab_coauthor['TRUE',]),2)`%) reported female leadership, suggesting that female ethnographers were more likely to report female leaders. However, after accounting for the non-independence of text records (i.e., multiple records from the same document) using a logistic mixed effects regression model, we found that although there was a positive effect of co-author gender on reporting female leadership (OR = `r signif(mm_coauthorOR, 2)`), it was not statistically significant (p = `r signif(mm_coauthorP, 2)`).
